Containers and R: the Rockerverse and beyond

### 2019-07-28 manuscript.Rmd
Update manuscript with suggested contents and responsibilities from #3

---
title: | 
       | Containers and R:
       | the Rockerverse and beyond
author:
  - name: Daniel Nüst
    affiliation: Institute for Geoinformatics, University of Münster 
    address:
    - Heisenbergstr. 2, 48149 Münster, Germany
    email: daniel.nuest@uni-muenster.de
abstract: >
  The Rocker project provides widely-used Docker images for R across different application scenarios. In this article we describe the downstream projects building upon Rocker and look beyond towards other container solutions and libraries to interact with them from R.
preamble: >
   \usepackage{longtable}
output: rticles::rjournal_article
---

@@

\citep{R}

### Containerization

### Rocker

## R images and containers

### Rocker

### Bioconductor

### MRO

### Renjin

### pqR

### Singularity

### OCI

## Packages

### Automation and packaging

- `containerit`
- `dockertest`
- `liftr`

### Control and provisioning

- `harbor`
- `docker`
- `RSelenium`
- `googleComputeEngineR`
- `analogsea`

## Downstream Applications

### Deployment and processing

- `plumber`
- `batchtools`

@@

### Workflows and reproducibility

### Scalable deployments

- Shiny talks @ useR!2017
- ...

>>>>

---
title: "Containers and R: the Rockerverse and beyond"
author:
  - name: Daniel Nüst
    affiliation: Institute for Geoinformatics, University of Münster 
    address:
    - Heisenbergstr. 2, 48149 Münster, Germany
    email: daniel.nuest@uni-muenster.de
  - name: Carl Boettiger
    #affiliation: -
    #email: -
  - name: Robrecht Cannoodt
    #affiliatin: -
    #email: -
  - name: Dirk Eddelbuettel
    #affiliation: -
    #email: -
  - name: Mark Edmondson
    #affiliation: -
    #email: -
  - name: Colin Fay
    #affiliation: -
    #email: -
  - name: Ben Marwick
    #affiliation: -
    #email: -
  - name: Noam Ross
    #affiliation: -
    #email: -
  - name: Nan Xiao
    #affiliation: -
    #email: - 
abstract: >
  The Rocker project provides widely-used Docker images for R across different application scenarios.
  This articles surveys describe downstream projects building upon Rocker.
  We also look beyond Rocker to other projects connecting containerisation with R.
  These use cases and the diversity of applications demonstrate the power of Rocker and containerisation for collaboration, effectivity, scalability, and transparency.
preamble: >
   \usepackage{longtable}
output: rticles::rjournal_article
---

@@

The \citep{R} community keeps growing: the number of new packages on CRAN keeps on rising, meetups, conferences, online courses and unconferences prosper, and the uptake in education and industry increases [REFs].
All this cements `R`'s role as the lingua franca of statistics, data visualisation, and computational research.
Part of this success can be traced back to the _Rocker Project_ ([https://www.rocker-project.org/](https://www.rocker-project.org/)).
The community project was introduced in 2014 and provides a number of [Docker](https://en.wikipedia.org/wiki/Docker_(software)) images for various use cases \citet{RJ-2017-065}.
The considerable uptake and continued evolution of _Rocker_'s suite has lead to numerous projects extending or building upon _Rocker_ images ranging from reproducible research to production deployments.
This article presents this _Rockerverse_ of projects.
It also introduces related activities connecting the R language and environment with other containerisation solutions.
The main contributions is a coherent picture of the current lay of the land of containers in, with, and for `R`.
This diversity includes demonstrators, early prototypes, and mature projects.

## Containerization and Rocker

_Do we need a generic intro here?_

## Images for alternative R distributions [@nuest]

- [MRO](https://github.com/nuest/mro-docker)
- [Renjin](https://github.com/nuest/renjin-docker)
- [pqR](https://github.com/nuest/pqr-docker)
- [FastR](https://github.com/nuest/fastr-docker)

## Other containerisation platforms [NN]

- R images for Singularity
- Running Rocker images with
  - Singularity
  - CoreOS rkt?
- nix?

## Non-Debian images [NN]

- Alpine images
- Images used by R-Hub (overlap with CI?)
- [Windows](https://github.com/nuest/rocker-win) [@nuest]
- https://github.com/jlisic/R-docker-centos

## Further Use-case Docker images [NN]

- part of Rocker (if not mentioned in \citet{RJ-2017-065}) ?
  - Binder
  - GPU [@noamross, @cboettig]
- [Bioconductor](http://bioconductor.org/help/docker/), https://github.com/Bioconductor/bioc_docker
- Gigantum stack
- ...

## Rocker for Research Compendia [@benmarwick]

## Packages for automation and packaging [@nuest]

- `containerit`
- `dockertest`
- `liftr` for OS-level reproducibility [@nanx]
- [`rize` for Shiny](https://github.com/cole-brokamp/rize)
- [`dockerfiler`](https://github.com/ColinFay/dockerfiler)

## Packages for control and provisioning [@nuest]

- [`stevedore`](https://github.com/richfitz/stevedore)
- [`babelwhale`](https://cran.r-project.org/web/packages/babelwhale/index.html): Running a Docker from R with Singularity or Docker as back-end. This is really useful in HPC environments where a user might not have root access but is able to install Singularity instead of Docker. [@rcannood]
- [`docker`](https://bhaskarvk.github.io/docker/)
- `RSelenium`
- [`googleComputeEngineR`](https://cloudyr.github.io/googleComputeEngineR/) (function `docker_run`)
- [`analogsea`](https://github.com/sckott/analogsea)
- [`harbor`](https://github.com/wch/harbor/)
- [`dockermachine`](https://github.com/cboettig/dockermachine)

## Continuous integration and continuous delivery [@noamross, @ColinFay]

- R-Hub
- DevOps
  - https://www.opencpu.org/posts/opencpu-with-docker/
- `r-ci`: https://github.com/ColinFay/r-ci
- [dynwrap](https://github.com/dynverse/dynwrap_containers/blob/master/.travis.yml) [@rcannood]
  - For this project, we use travis-ci to build rocker-derived containers, test them, and only push them to docker hub (from travis-ci.org) if the integration tests succeed.
- ...

## Common work environments [@noamross]

- ...
- in education
  - ready to use learning environments
  - [`r-db`](https://github.com/ColinFay/r-db) [@ColinFay]
- [RCloud](https://github.com/att/rcloud/tree/master/docker) ?

## Deployment, processing, cloud

- Docker images for cloud services [@MarkEdmondson1234]
- [`plumber`](https://www.rplumber.io/docs/hosting.html#docker)
- `batchtools` \citep{Lang2017batchtools} can [schedule jobs with Docker Swarm](https://mllg.github.io/batchtools/reference/makeClusterFunctionsDocker.html)
- scalable deployments, e.g. start with numerous Shiny talks mentioning Rocker at useR!2017
- [dynmethods](https://github.com/dynverse/dynmethods) [@rcannood]: In order to evaluate ±50 computational methods which all used different environments (R, Python, C++, ...), we wrapped each of them in a docker container and can execute these methods from R. Again, all of these containers are being built on travis-ci, and will only be pushed to docker hub if the integration test succeeds.
- [ShinyProxy](https://www.shinyproxy.io/) [creates a container](https://github.com/openanalytics/shinyproxy/blob/master/src/main/java/eu/openanalytics/services/DockerService.java#L388) for each user

## R versions [@ColinFay]

- run code with Versions of R: https://srv.colinfay.me/r-online
- old R versions: https://github.com/rocker-org/rocker-versioned/issues/138
- [semantic version tags for Rocker](https://github.com/rocker-org/rocker-versioned/#version-tags)
- debugging memory problems across multiple versions: https://github.com/wch/r-debug

@@

## Discussion/outlook [all, please add bullet points]

- Missing pieces?
- consolidation (e.g. via packages using `dockerfiler` and `stevedore`)
- Common themes
  - reproducibility
- will knowledge about containers continue to spread?
- what is needed for even more containers with R?
- ...
